Supermartingales and Termination Analysis of Statistical PCF

### 2020-10-27 probabilisticProgrammingMartingales.tex
Prove Lemma 1 (downAcrossUp).
The proof turned out to be more complicated than expected, because
it's non-trivial to show that the sequence of rearrangements required
actually terminates. It was possible to adapt a proof of the Church-
Rosser theorem to prove termination.

\begin{lemma} \label{downAcrossUp}
If $(N_1,\alpha_1) \sim^* (N_2,\alpha_2)$, there is some descendants $N'_1, N'_2$ of $N_1$ resp.~$N_2$, and a position $\alpha'$ in both of them such that $(N_1,\alpha_1) \sim_p^* (N'_1, \alpha') \sim_c^* (N'_2, \alpha') \sim_p^* (N_2, \alpha_2)$.
\end{lemma}
\begin{proof}
The $\sim$ relation can be split into $\sim_\downarrow \cup \sim_\uparrow$, where $(A,\alpha) \sim_\downarrow (B,\beta)$ if either $A \to B$ and $(A,\alpha) \sim_p (B,\beta)$ or $(A,\alpha) \sim_c (B,\beta)$, and similarly $(A,\alpha) \sim_\uparrow (B,\beta)$ if either $B \to A$ and $(A,\alpha) \sim_p (B,\beta)$ or $(A,\alpha) \sim_c (B,\beta)$. The sequence of $\sim$ relations linking $N_1$ to $N_2$ can then be expressed as an alternating sequence of $\sim_\downarrow^*$ and $\sim_\uparrow^*$s, starting with $\sim_\downarrow^*$ and ending with $\sim_\uparrow^*$, and the proof proceeds by induction on the number of these alternations.

In the base case, that $(N_1,\alpha_1) \sim_\downarrow^* X \sim_\uparrow^* (N_2,\alpha_2)$, \dots

\akr{Actually proving that this process terminates will be a pain although I'm confident it will work, but the basic idea is that the sequence of $\sim$ steps can be altered by repeatedly replacing the subsequence $\sim_c, \sim_{p,\downarrow}$ with $\sim_{p,\downarrow}, \sim_c$, the subsequence $\sim_{p,\uparrow}, \sim_c$ with $\sim_c, \sim_{p,\uparrow}$, and the subsequence $\sim_{p,\uparrow}, \sim_{p,\downarrow}$ with $\sim_{p,\downarrow}^*, \sim_c, \sim_{p,\uparrow}^*$, until everything is in the correct order.}
\end{proof}

>>>>

\begin{lemma} \label{downAcrossUp}
If $(N_1,\alpha_1) \sim^* (N_2,\alpha_2)$, there is some descendants $N'_1, N'_2$ of $N_1$ resp.~$N_2$, and a position $\alpha'$ in both of them such that $(N_1,\alpha_1) \sim_p^* (N'_1, \alpha') \sim_c^* (N'_2, \alpha') \sim_p^* (N_2, \alpha_2)$.
\end{lemma}
\begin{proof}
  The $\sim_p$ relation can be split into $\sim_\downarrow \cup \sim_\uparrow$, where $(A,\alpha) \sim_\downarrow (B,\beta)$ if $A \to B$ and $(A,\alpha) \sim_p (B,\beta)$, and similarly $(A,\alpha) \sim_\uparrow (B,\beta)$ if $B \to A$ and $(A,\alpha) \sim_p (B,\beta)$. At each stage of this proof, it will be assumed that the $\sim_c$ steps and the $\sim_p$ steps in the sequence from $(N_1,\alpha_1)$ to $(N_2,\alpha_2)$ are rearranged such that there is never a $\sim_c$ immediately before a $\sim_\downarrow$, and there is never a $\sim_c$ immediately after a $\sim_\uparrow$. This rearrangement is always possible, because if there is some subsequence $(A,\alpha) \sim_c (B,\alpha) \sim_\downarrow (C,\beta)$, then there is an alternate path $(A,\alpha) \sim_\downarrow (B',\beta) \sim_c (C,\beta)$, where the reduction and the $\sim_\downarrow$ from $A$ to $B'$ are the same as those from $B$ to $C$, and the $\sim_c$ step is the same, but with the reduction sequences $O_1 \to^* O_1'$ and $O_2 \to^* O_2'$ in the definition of $\sim_c$ extended by the reduction $B \to C$. With these rearrangements assumed, it is sufficient to prove that the $\sim_p$ steps can be rearranged so that all of the $\sim_\downarrow$ steps come before all of the $\sim_\uparrow$ steps (possibly introducing some $\sim_c$ steps in the process).

  The rearrangement to put all of the $\sim_\downarrow$s before all of the $\sim_\uparrow$s proceeds by induction on the number of $\sim_\downarrow$s that occur after the first occurrence of $\sim_\uparrow$. If it's 0, all of the $\sim_p$s are already in the correct order, and we're done. Otherwise, take the subsequence from the first $\sim_\uparrow$ to the first $\sim_\downarrow$ after that. This must be rearranged to some $\sim_\downarrow$s followed by some $\sim_\uparrow$s, then once that's done, the number of $\sim_\downarrow$s after the first $\sim_\uparrow$ in the overall sequence will have decreased by 1.

  Let this subsequence be $A \sim_\uparrow^n B \sim_\downarrow C$. Suppose for induction that $A \sim_\uparrow^k (D,\delta) \sim_\downarrow^* (E,\epsilon) \sim_\uparrow^* C$ for some $D, \delta, E, \epsilon$, and $0 \leq k \leq n$, where in the reduction sequence $D \to^* E$, for any pair of reductions at positions which aren't disjoint, the reduction at the innermost (greater) position occurs first. Reduction sequences with this property will be called ``parallel'', as it is analagous to parallel reduction as in \akr{TODO: find this reference, for the proof of the Church-Rosser theorem by Tait and Martin-L\"of}. This induction is in reverse, with $k$ decreasing from $n$ to $0$. If $k = n$, simply set $(D,\delta) = B$ and $(E,\epsilon) = C$. If $k = 0$, then $A$ is related to $B$ by a sequence of $\sim_\downarrow$s then $\sim_\uparrow$s, as desired. In the intermediate steps, $k$ must be decreased by one, so a subsequence of one $\sim_\uparrow$ followed by a parallel sequence of $\sim_\downarrow$s must be replaced by a parallel sequence of $\sim_\downarrow$s followed by some $\sim_\uparrow$s.

  For any individual $\sim_\uparrow$ then $\sim_\downarrow$ pair, let $(A,\alpha) \sim_\uparrow (B,\beta) \sim_\downarrow (C,\gamma)$. The skeleton $B$ reduces to both $A$ and $C$. If these are the same reduction, then $(A,\alpha) = (C,\gamma)$ and this subsequence may be removed. If they are different, then the way they overlap corresponds to one of the cases in the definition of $\sim_c$, either case a if the reduction positions are disjoint, or one of the cases b--f if one of the reduction positions is inside the other. In any case, there is a reduction sequence from each of $A$ and $C$ to some common skeleton $D$ ($O_1$ and $O_2$ in the definition of $\sim_c$). For the same reason that $(B,\beta) \sim_\downarrow (C,\gamma)$, there is some $\delta$ such that $(A,\alpha) \sim_\downarrow^* (D,\delta)$, and similarly for the same reason that $(B,\beta) \sim_\downarrow (A,\alpha)$, for the same $\delta$, $(C,\gamma) \sim_\downarrow^* (D,\delta)$. There are a lot of cases to check for this statement, but all of them are rather simple, and similar to each other. There is therefore an alternative $\sim^*$ sequence from $(A,\alpha)$ to $(C,\gamma)$, of the form $(A,\alpha) \sim_\downarrow^* (D_1,\delta) \sim_c (D_2,\delta) \sim_\uparrow^* (C,\gamma)$, where $D_1$ and $D_2$ are the alternative reduction sequences leading to the same skeleton $D$. There are only multiple $\sim_\downarrow$ steps in the result if the position of the reduction $B \to C$ is inside the position of the reduction $B \to A$, and similarly, there are only multiple $\sim_\uparrow$ steps in the result if the position of the reduction $B \to A$ is inside the position of the reduction $B \to C$. If there are multiple $\sim_\downarrow$ steps, they are not necessarily parallel, but in the only case where they aren't (case f of $sim_c$), the subsequence of $\sim_\downarrow$ steps with the reductions at $\alpha;\lambda;@_1;\beta$ then $\alpha;\lambda;@_1;\gamma_i;\tY;\lambda;\beta$ for each $i$ where $\gamma_i \geq \beta$ (in the notation of the definition of $\sim_c$) can be replaced by $\sim_\downarrow$s with the reductions at $\alpha;\lambda;@_1;\gamma'_i;\tY;\lambda;\beta$ then $\alpha;\lambda;@_1;\beta$, then another $\sim_c$, where $(\gamma'_i)$ is the list of positions in $A$ (\emph{before} the reduction at $\beta$ to $A'$, unlike $(\gamma_i)$) where $A | \gamma'_i = x$ and $\gamma'_i > \beta$.

  The effect of this rearrangement is (ignoring the $\sim_c$s) to swap the order of the $\sim_\uparrow$ and the $\sim_\downarrow$, except that if the positions of one of these reductions is inside the other, that inner $\sim_p$ may be duplicated. If the positions are disjoint, both of them are unchanged (corresponding to case a of $\sim_c$), and if one is inside the other, then the outer position is unchanged and the other resultant positions, although they aren't equal to the original inner position, are still inside the outer position. Taking the sub-sequence of one $\sim_\uparrow$ followed by a parallel sequence of $\sim_\downarrow$s, this rearrangement can be repeatedly applied to move the $\sim_\uparrow$ further along the sequence. If at some point a $\sim_\uparrow$ and $\sim_\downarrow$ match (by relating the same (skeleton, position) pairs in opposite directions), they may be removed and the process may stop early. The $\sim_\uparrow$ may be duplicated if it passes a $\sim_\downarrow$ whose position is outside of the $\sim_\uparrow$'s position, but by the assumption that the sequence of $\sim_\downarrow$s is initially parallel, no position inside of this occurs later in the sequence, therefore all of the resultant $\sim_\uparrow$s pass the remaining $\sim_\downarrow$s without changing or duplicating them. The $\sim_\uparrow$s may be further duplicated, but the number of steps left in this process is bounded by the product, for all of the remaining $\sim_\downarrow$s, of $2 + \text{the number of occurrences of the variable relevant to the reduction}$, because for each $\sim_\downarrow$, that is more than the number of duplicates it could produce for any $\sim_\uparrow$ that passes it. It is also possible (earlier) for some of the $\sim_\downarrow$s to be duplicated, but the only case where this process would not terminate, that both the $\sim_\downarrow$s and the $\sim_\uparrow$s continue being duplicated forever so that the number of switches left to do never decreases, is prevented by the parallelness condition, as all of the duplications of $\sim_\downarrow$s occur before all of the duplications of $\sim_\uparrow$s.

  It still remains to be shown that the sequence of $\sim_\downarrow$s left at the end of this can be made parallel. The changes that may have occurred since the previous version of this sequence (which was known to be parrallel already), are that some of the $\sim_\downarrow$s may have been duplicated by passing the $\sim_\uparrow$, and if that $\sim_\uparrow$ matches one of the $\sim_\downarrow$s, that $\sim_\downarrow$ is removed. The position of the $\sim_\uparrow$'s reduction is the same for all of these duplications, because it is not changed until later in the sequence, when the $\sim_\downarrow$s with positions outside of that may occur. The resultant positions after a duplication are in the same position relative to each other, and therefore the order of the remaining $\sim_\downarrow$s is automatically parallel, except that (letting the position of the $\sim_\uparrow$'s reduction be $\alpha$) if there are $\sim_\downarrow$s with reductions at positions $\alpha;@_1;\lambda;\beta$ and $\alpha;@_2;\gamma$, or $\alpha;\tY;\lambda;\beta$ and $\alpha;\tY;\lambda;\gamma$, some of the resultant $\sim_\downarrow$s' positions may be inside each other where they were originally not. Similarly to the case where a $\tY$ reduction and a reduction at a position inside it are swapped though, the order can be fixed by swapping some of the $\sim_\downarrow$ steps with each other. In the first case, the reduction at position $\alpha;@_1;\lambda;\beta$ ends up at $\alpha;\beta$, and the reduction at position $\alpha;@_2;\gamma$ ends up at the positions $\alpha;\delta_i;\gamma$, where $(\delta_i)$ is the positions of the variable relevant to the $\sim_\uparrow$'s reduction inside its lambda. The positions of the variable in the lambda may be different before the reduction at $\beta$, but the positions of the redexes corresponding to the original redex at $\alpha;@_2;\gamma$ are changed similarly. Rather than reducing at $\alpha;\beta$ then at $\alpha;\delta_i;\gamma$, it is therefore possible to reach the same point by reducing at $\alpha;\delta'_i;\gamma$ then at $\alpha;\beta$, where $(\delta'_i)$ is the positions of the relevant variable within the lambda before the reduction at $\alpha;@_1;\lambda;\beta$ (and again, a $\sim_c$ of some sort must also be introduced). The other case, that the $\sim_\uparrow$'s reduction is a $\tY$-reduction, is similar. Each reduction which is duplicated has one image at $\alpha;\lambda;@_1;$its original relative position, and one for each occurrence of the variable. Those corresponding to variable occurrences may need to be moved to before those at $\lambda;@_1$.

  In summary, to reach the desired order of $\sim_p$ steps, each $\sim_\downarrow$ may have to be moved past some $\sim_\uparrow$ steps, and it may get duplicated in the process, but the resultant sequence of $\sim_\downarrow$s can all be put in parallel order, and it has been shown that moving a $\sim_\uparrow$ past a sequence of $\sim_\downarrow$s is always possible if they're in parallel order, therefore the overall process terminates and reaches a point where all of the $\sim_\downarrow$s precede all of the $\sim_\uparrow$s, and all of the $\sim_c$s are in between them. All of these rearrangements leave the end points of the sequence unchanged, therefore this is an alternative sequence of $\sim$ steps between $(N_1,\alpha_1)$ and $(N_2,\alpha_2)$ of the form $(N_1,\alpha_1) \sim_\downarrow^* (N_1', \alpha') \sim_c^* (N_2', \alpha') \sim_\uparrow^* (N_2,\alpha_2)$.
\end{proof}
